A Linear Technique to Predict Non-Normal Turbulence Applied to a Magnetized Plasma

### 2013-12-20 Transient_growth_prl.tex
Dec 20 2

\begin{figure}
\centerline{\includegraphics[width=0.4\textwidth]{transient_n0_n1_growth}}
\caption{Linear evolution of energy starting from a turbulent initial state. The $n=0$ curves have an initial period of transient growth before exponentially decaying.}
\label{transient_n0_n1_growth}
\end{figure}

This convective transport of density filaments is akin to the paradigmatic ``lift-up'' mechanism in hydrodynamic shear flows whereby streamwise vortices drive streamwise streaks~\cite{krommes1999}.
Both are transient growth processes. We see this in our simulations by following the evolution of the energy of several $m,n$ modes after turning off the nonlinearities in an already
turbulent simulation. A few representative modes are shown in Fig.~\ref{transient_n0_n1_growth}. The linear transient growth of the filamentary $n=0$ structures is evident as their modes 
grow transiently before decaying exponentially at the rate of their least stable eigenmode. Such behavior is rather unituitive to some because all $n=0$ linear eigenmodes are stable, 
and the normal mode paradigm has conditioned us to believe that a linear superposition of such eigenmodes should decay. But as the hydrodynamics community discovered, linear
superpositions of nonorthogonal vectors may behave unintuitively.
Notice now that the $n=1,m=20$ mode decays transiently before growing exponentially with growth rate of the most unstable eigenmode. This isn't that interesting since this behavior
can occur in normal systems, but it nevertheless explains later results.

\begin{figure}
\centerline{\includegraphics[width=0.45\textwidth]{spec_nl_gamma}}
\caption{Linear and turbulent growth rate spectra for $n=0$ (solid lines) and $n=1$ (dashed lines) Fourier components. The linear growth rates are those of the least stable eigenmodes,
while the turbulent growth rates represent $\pdiff{E}{t}|_{\rm{lin}}/2 E$ from the nonlinear simulation. The shaded region marks the $1 \sigma$ spread in the turbulent spectrum,
obtained from the distribution of growth rates in the nonlinear simulation.}
\label{spec_nl_gamma}
\end{figure}

Since the transient growth, which is behind subcritical turbulence, is a purely linear phenomenon, it has long been a goal of researchers to predict the onset of subcritical turbulence using
only linear, non-modal calculations. It is our goal, however, to use linear calculations to understand and even predict turbulent properties.
Perhaps one of the most useful statistical properties that we can predict is the turbulent growth rate spectrum.
To obtain a growth rate from the nonlinear simulation, we begin with the expression $\diff{E(m,n)}{t}\big|_{\rm{lin}} = Q(m,n) + D(m,n)$, 
which accounts for the total net energy injection at each $m,n$ into the fluctuations. 
An effective instantaneous growth rate $\gamma_t(m,n) = \diff{E(m,n)}{t}\big|_{\rm{lin}}/ 2 E(m,n)$ can then be defined. 
This effective growth rate is calculated from the spatial structure of the fluctuations, so it is well-defined for both turbulent fluctuations and eigenmode fluctuations. If calculated
from a linear simulation run for a sufficiently long time, $\gamma_t(m,n)$ is the same as the eigenmode growth rate of the least stable linear eigenmode for each $m,n$.
For turbulence, $\gamma_t(m,n)$ is generally not equal to an eigenmode growth rate, especially in the case where the system is non-normal.
We plot this growth rate in Fig.~\ref{spec_nl_gamma} for $n=0$ and $n=1$ components. The turbulent growth rate curves are calculated from this formula, 
averaged over a time of about $3 \times 10^{4} \ \omega_{ci}^{-1}$ during the saturated turbulent phase of the nonlinear simulation.
Also, we show the $1 \sigma$ contour, which indicates how greatly $\gamma_t$ varies over time.
Fig.~\ref{spec_nl_gamma} indicates that while the linear eigenmodes are unstable for $n=1$ and stable for $n=0$,
the turbulent growth rates are positive for $n=0, m<50$ (subcritical-like behavior), but negative elsewhere. 
Fig.~\ref{transient_n0_n1_growth} hinted at these characteristics, but this figure represents quantitative confirmation.

Our goal now is to reproduce the turbulent growth rate curves in Fig.~\ref{spec_nl_gamma} without performing nonlinear simulations. This is difficult because the growth rates depend on turbulent
structures, which for non-normal systems, may be very different from eigenmode structures, and which change in time due to transient evolution.
It is clear, though, that the turbulent structures are a result of both linear evolution and nonlinear energy transfer. 
If we imagine that the nonlinearities randomize the turbulent structures at each wavenumber while the linearities evolve the structures
deterministically, we can develop a procedure for choosing proper spatial structures to use in calculating effective growth rates.
To elucidate this procedure, we begin by taking Eqs.~\ref{ni_eq}-\ref{te_eq} and Fourier decomposing in the azimuthal and axial directions.
Then, we discritize in the radial direction and approximate radial derivatives with finite differences. 
The resulting system of equations may be written in matrix form:

@@

\begin{figure}
\centerline{\includegraphics[width=0.5\textwidth]{m20n0_time_evs}}
\caption{a) The maximum growth ratio curve $G_{\rm{max}}(t)$ (dashed line) for $n=0,m=20$, and an ensemble of growth ratio curves (solid gray lines, with the solid black line the ensemble average)
that start with a random initial condition $u(0)$ and evolve under the linear operator. b) The ensemble curves with $t=\omega_s^{-1}$ indicated, at which we cut off the evolution.}
\label{m20n0_time_evs}
\end{figure}

>>>>

\begin{figure}
\centerline{\includegraphics[width=0.55\textwidth]{n0_n1_growth}}
\caption{a) Linear evolution of energy starting from a turbulent initial state. The $n=0$ curves have an initial period of transient growth before exponentially decaying.
b) Linear and turbulent growth rate spectra for $n=0$ (solid lines) and $n=1$ (dashed lines) Fourier components. The linear growth rates are those of the least stable eigenmodes,
while the turbulent growth rates represent $\pdiff{E}{t}|_{\rm{lin}}/2 E$ from the nonlinear simulation. The shaded region marks the $1 \sigma$ spread in the turbulent spectrum,
obtained from the distribution of growth rates in the nonlinear simulation.}
\label{n0_n1_growth}
\end{figure}

This convective transport of density filaments is akin to the paradigmatic ``lift-up'' mechanism in hydrodynamic shear flows whereby streamwise vortices drive streamwise streaks~\cite{krommes1999}.
Both are transient growth processes. We see this in our simulations by following the evolution of the energy of several $m,n$ modes after turning off the nonlinearities in an already
turbulent simulation. A few representative modes are shown in Fig.~\ref{n0_n1_growth} a). The linear transient growth of the filamentary $n=0$ structures is evident as their modes 
grow transiently before decaying exponentially at the rate of their least stable eigenmode. Such behavior is rather unituitive to some because all $n=0$ linear eigenmodes are stable, 
and the normal mode paradigm has conditioned us to believe that a linear superposition of such eigenmodes should decay. But as the hydrodynamics community discovered, linear
superpositions of nonorthogonal vectors may behave unintuitively.
Notice now that the $n=1,m=20$ mode decays transiently before growing exponentially with growth rate of the most unstable eigenmode. This isn't that interesting since this behavior
can occur in normal systems, but it nevertheless explains later results.

Since the transient growth, which is behind subcritical turbulence, is a purely linear phenomenon, it has long been a goal of researchers to predict the onset of subcritical turbulence using
only linear, non-modal calculations. It is our goal, however, to use linear calculations to understand and even predict turbulent properties.
Perhaps one of the most useful statistical properties that we can predict is the turbulent growth rate spectrum.
To obtain a growth rate from the nonlinear simulation, we begin with the expression $\diff{E(m,n)}{t}\big|_{\rm{lin}} = Q(m,n) + D(m,n)$, 
which accounts for the total net energy injection at each $m,n$ into the fluctuations. 
An effective instantaneous growth rate $\gamma_t(m,n) = \diff{E(m,n)}{t}\big|_{\rm{lin}}/ 2 E(m,n)$ can then be defined. 
This effective growth rate is calculated from the spatial structure of the fluctuations, so it is well-defined for both turbulent fluctuations and eigenmode fluctuations. If calculated
from a linear simulation run for a sufficiently long time, $\gamma_t(m,n)$ is the same as the eigenmode growth rate of the least stable linear eigenmode for each $m,n$.
For turbulence, $\gamma_t(m,n)$ is generally not equal to an eigenmode growth rate, especially in the case where the system is non-normal.
We plot this growth rate in Fig.~\ref{n0_n1_growth} b) for $n=0$ and $n=1$ components. The turbulent growth rate curves are calculated from this formula, 
averaged over a time of about $3 \times 10^{4} \ \omega_{ci}^{-1}$ during the saturated turbulent phase of the nonlinear simulation.
Also, we show the $1 \sigma$ contour, which indicates how greatly $\gamma_t$ varies over time.
Fig.~\ref{n0_n1_growth} b) indicates that while the linear eigenmodes are unstable for $n=1$ and stable for $n=0$,
the turbulent growth rates are positive for $n=0, m<50$ (subcritical-like behavior), but negative elsewhere. 
Fig.~\ref{n0_n1_growth} a) hinted at these characteristics, but this figure represents quantitative confirmation.

Our goal now is to reproduce the turbulent growth rate curves in Fig.~\ref{n0_n1_growth} b) without performing nonlinear simulations. This is difficult because the growth rates depend on turbulent
structures, which for non-normal systems, may be very different from eigenmode structures, and which change in time due to transient evolution.
It is clear, though, that the turbulent structures are a result of both linear evolution and nonlinear energy transfer. 
If we imagine that the nonlinearities randomize the turbulent structures at each wavenumber while the linearities evolve the structures
deterministically, we can develop a procedure for choosing proper spatial structures to use in calculating effective growth rates.
To elucidate this procedure, we begin by taking Eqs.~\ref{ni_eq}-\ref{te_eq} and Fourier decomposing in the azimuthal and axial directions.
Then, we discritize in the radial direction and approximate radial derivatives with finite differences. 
The resulting system of equations may be written in matrix form:

@@

\begin{figure}
\centerline{\includegraphics[width=0.55\textwidth]{m20n0_time_evs}}
\caption{a) The maximum growth ratio curve $G_{\rm{max}}(t)$ (dashed line) for $n=0,m=20$, and an ensemble of growth ratio curves (solid gray lines, with the solid black line the ensemble average)
that start with a random initial condition $u(0)$ and evolve under the linear operator. b) The ensemble curves with $t=\omega_s^{-1}$ indicated, at which we cut off the evolution.}
\label{m20n0_time_evs}
\end{figure}
